---
title: "Blog Post 4"
subtitle: "Continuing Exploratory Analysis "
author: ""
date: "2024-04-03"
date-modified: "2024-04-03"
draft: FALSE
---
```{r}
library(tidyverse)
library(MASS)

compas_scores_clean <- read_csv("dataset/compas-scores-clean.csv")

model <- polr(as.factor(DecileScore) ~ Sex, data = compas_scores_clean, Hess=TRUE)
model
summary(model)
predictions <- predict(model, compas_scores_clean, type="class")
ggplot(compas_scores_clean, aes(x=Sex, y=DecileScore)) +
  geom_boxplot() +
  labs(title="Distribution of DecileScores by Sex",
       x="Sex",
       y="DecileScore") +
  theme_minimal()
```
Both Female and Male categories have median scores (indicated by the line inside the box) that are approximately at the same level, suggesting that the central tendency of scores is similar across both sexes.The range for Males is slightly larger than for Females, which suggests that scores for Males are more spread out. There do not appear to be any outliers indicated for either Females or Males, suggesting that all the scores fall within a relatively expected range. The male has wider IQR than Female’s. Also, both graphs are skewed to the right.
=======
## Relationship Between Race and Decile Score

```{r}

df <- compas_scores_clean %>% 
  mutate(AA = if_else(Race == "African-American" | Race == "African-Am", 1, 0))

race_model <- lm(DecileScore ~ AA, data = df)
summary(race_model)
predictions <- predict(race_model, df)

ggplot(compas_scores_clean, aes(x=Race, y=DecileScore)) +
  geom_boxplot() +
  labs(title="Distribution of DecileScores by Race",
       x="Race",
       y="DecileScore") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
To explore the relationship between race and decile scores, we decided to use linear regression analysis. The results of the regression model revealed a statistically significant correlation where being classified as African-American was associated with an increase in the decile score by approximately 1.3 points. However, the model’s adjusted R-squared value was 0.06062, indicating that race accounts for only about 6% of the variation in decile scores, suggesting other factors also play a significant role. The accompanying boxplot clearly illustrates higher median scores for African-Americans, indicating a potential systemic trend that would be worth investigating further.






